Assessing linkages between watershed nutrient loading and estuary water quality in Lavaca Bay, Texas

### 2023-02-11 Schramm-Manuscript-2023/Schramm-Manuscript-2023.Rmd
edits. Add IT approach to methods, add lagged responses in discussion.

<!---
Need to add model selection criteria!
--->

@@

Based on criteria provided by @moriasiHydrologicWaterQuality2015, GAMs ranged
from "satisfactory" to "very good" using median NSE, r^2^ and PBIAS 
calculated using repeated 5-fold cross validation on predicted and measured
nutrient loads. NO~3~ GAM models had median NSE values of 0.34 and 0.48 at
USGS-08164000 and USGS-08164390 respectively. Median r^2^ values were 0.70
(USGS-08164000) and 0.87 (USGS-08164525) and PBIAS values were 2.00
(USGS-08164000) and 10.90 (USGS-08164525) for NO~3~ loads. Median NSE values were
0.80 (USGS-08164000) and 0.91 (USGS-08164525) for TP loads. Median r^2^ values 
for TP loads were 0.93 (USGS-08164000) and 0.99 (USGS-08164525) and median PBIAS 
values were -7.20 (USGS-08164000) and -3.30 (USGS-08164525). Density plots of
metrics show similar distribution of values between sites for the 
same parameter, with the exception r^2^ values for NO~3~ loads where 
USGS-08164000 showed a much larger variance in values compared to USGS-08164525
(Figure \ref{fig:fig2}). TP GAMS had higher average NSE and r^2^ values and 
lower variance in metric values compared to NO~3~.

@@

Annual NO~3~ and TP loads show considerable variation, generally following
patterns in discharge (Figure \ref{fig:fig3}, Figure \ref{fig:fig4}). 
Flow-normalized TP loads at both sites and flow-normalized loads at 
USGS-08164000 indicate watershed based loads have not changed much over time
when accounting for changes in streamflow (Figure \ref{fig:fig3}). 
Flow-normalized loads at USGS-08164000 show small variation over time with 
some decreases in NO~3~ loads since 2013. Aggregated across both sites, the mean 
annual NO~3~ load 2005 through 2020 was `r mean_no3` kg (`r mean_no3_low` kg - 
`r mean_no3_upper` kg, 90% CI). Annual NO~3~ loads ranged from `r min_no3` kg 
in `r min_no3_year` to `r max_no3` kg in `r max_no3_year`.

Total annual TP loads ranged from `r min_tp` kg in `r min_tp_year` to 
`r max_tp` kg  in `r max_tp_year`. Mean annual TP loading from 2005 through 2020
was `r mean_tp` kg (`r mean_tp_low` kg - `r mean_tp_upper` kg, 90% CI).
On average, USGS-08164525 accounted for `r no3_mean_prop` of NO~3~ loads and 
`r tp_mean_prop` of TP loads from 2005 through 2020. However, during periods of extreme 
drought the Lavaca River (USGS-08164000) became the primary source of 
nutrient loading in the watershed with the Navidad River only accounting for 
`r min_no3_prop` and `r min_tp_prop` of NO~3~ and TP loads in 2011 (Figure 
\ref{fig:fig4}).

@@

Other studies have published estimates of TP loading estimates for the Lavaca
River (USGS-08164000) over various time periods (Table \ref{tab:tpcomp}).
Converted to average annual yield, the estimates of TP loading in this study
are within the ranges of previous studies [@dunnTrendsNutrientInflows1996; @rebich_sources_2011; @omaniEstimationSedimentNutrient2014; @wise_spatially_2019], with the TP 
estimates in @dunnTrendsNutrientInflows1996 being notably lower. Given that none
of the studies identify substantially sized trends in TP, it is is possible that 
the period used in @dunnTrendsNutrientInflows1996 was drier on average than the
other studies. The SPARROW models used in @rebich_sources_2011 and 
@wise_spatially_2019 utilize a version of LOADEST in the underlying load
estimation procedure, so a difference due to methodology alone is unlikely.

@@

Cross-validation of the GAM loading models indicated that GAMs performed well on
average at predicting daily nutrient loading values. The variance in scores was
very high indicating subsets of values were problematic at characterizing functional
relationships between nutrients and predictors. All of the water
quality data for the two river sites in the TCEQ database were ambient water
quality data, collected to be representative of typical conditions. This results
in few data collected at the highest portions of the flow duration curve
under which the majority of loadings take place. Supplementary flow-biased 
monitoring targeting storm- or high-flow conditions is recommended here to
improve the precision of GAM predictions [@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017].

The non-linear temporal water quality trends identified using GAMs differed
slightly from the trends identified by @bugica_water_2020. This is
not unexpected due to the different time periods, different methodology, and
generally small slopes identified for most of the significant water quality
parameters in prior work. The trend in DO and 
cholorophyll-*a* concentrations are stable in comparison to other Texas
estuaries that might be facing larger demands for freshwater diversions, higher
population growth, and more intense agricultural production 
[@wetzWaterQualityDynamics2016; @bugica_water_2020]. The trend of 
increasing NO~*x*~ concentration at the upper-bay TCEQ-13563 site and recent 
increases in TKN concentration at the mid-bay TCEQ-13383
site are concerning due to the nitrogen limitation identified in many Texas 
estuaries [@gardnerNitrogenFixationDissimilatory2006; 
@houTransformationFateNitrate2012; @doradoUnderstandingInteractionsFreshwater2015; 
@paudelRelationshipSuspendedSolids2019; @wetz_exceptionally_2017] 
and the relatively low ambient concentrations observed in Lavaca Bay.

The strong positive effect of freshwater inflow on NO~*x*~, TKN, and TP are
suggestive of nonpoint watershed sources, consistent with watershed uses, and
consistent with other studies relating freshwater inflow with nutrient 
concentrations in Lavaca Bay and other estuaries [@russell_effect_2006; 
@caffreyHighNutrientPulses2007; @peierlsNonmonotonicResponsesPhytoplankton2012; 
@palmerImpactsDroughtsLow2015; @ciraPhytoplanktonDynamicsLowInflow2021]. Inflow 
had a non-linear relationship with TKN at the two upstream sites, with TKN 
increasing as freshwater inflow transitioned from low to moderate levels. 
At higher freshwater inflows, the effect was attenuated, possibly indicating a 
flushing effect at higher freshwater inflow. No relationship between TKN and 
freshwater inflow were observed at TCEQ-13384 located in the lower reach of 
Lavaca Bay. Tidal flushing from Matagorda Bay might be responsible for 
decreasing TKN concentrations and limiting the effects of freshwater inflow in 
lower reaches of Lavaca Bay. @russell_effect_2006 suggested the processing of 
organic loads in the upper portions of Lavaca Bay reduces the transport of 
nutrients into the lower reaches of the Bay.

Freshwater inflow also had a strong positive effect on chlorophyll-*a* at 
the upper- and mid-bay sites. The upper-bay site, TCEQ-13563, showed decreases
in chlorophyll-*a* at the highest freshwater inflow volumes. Freshwater flushing
or increases in turbidity are associated with decreases in chlorophyll-*a* in
other estuaries [@peierlsNonmonotonicResponsesPhytoplankton2012; 
@cloernPhytoplanktonPrimaryProduction2014]. No relationships between inorganic
nitrogen or TP loadings with chlorophyll-*a* were observed. Due to the lack
of TKN loading information, no assessment between organic nitrogen loads and
chlorophyll-*a* were possible.

Although other studies have identified complex relationships between estuary 
nutrient concentrations, nutrient loading and chlorophyll-*a* concentrations in 
Texas estuaries [@ornolfsdottirNutrientPulsingRegulator2004; 
@doradoUnderstandingInteractionsFreshwater2015; @ciraPhytoplanktonDynamicsLowInflow2021; @tominackVariabilityPhytoplanktonBiomass2022], 
this study specifically used flow-adjusted freshwater derived nutrient loads to 
parse out contributions from changes in nutrient loadings while accounting for 
variations in load due to flow. Loading GAMs indicated no evidence of changes in 
flow-normalized TP loads in either river, and no changes in flow-normalized NO~3~
loads in the Navidad River. The small changes in flow-normalized NO~3~ loads in
the Lavaca River are probably masked under most conditions by discharge from the
Navidad River. Given the relatively little variation in flow-normalized loads,
it can be expected that they would contribute little to the variance in downstream
water quality.

GAMs did not identify responses in DO concentration to inflows or nutrient loads.
The seasonality term in the temporal GAM models explained a substantial amount of DO
variation at all of the sites. Responses of estuary metabolic processes and 
resulting DO concentrations can be quite complicated and often locally specific [@caffreyFactorsControllingNet2004]. While the lack of total nitrogen or TKN 
loading data hinders interpretation, the large seasonal effect on DO suggests 
physical factors play an important role and should be included in future models.
Prior work suggests that Lavaca Bay may not be limited by nutrients alone, 
with high turbidity or nutrient processing in upper portions of the Bay or 
intertidal river limiting production [@russell_effect_2006].

@@

GAM models appear to provide reliable estimates of nutrient loads in the Lavaca
Bay watershed. However, additional flow-biased data collection efforts would
decrease the prediction variance and probably improve accuracy at critical 
high flow events. Ongoing projects will fill data gaps for total nitrogen and 
TKN loading. This study, consistent with others along the Texas coast, found
strong effects of freshwater flow on nutrient and chlorophyll-*a* 
concentrations. DO concentrations, dominated by seasonal effects, did not show
strong direct responses to freshwater flow. Small variance in flow-adjusted
nutrient loads indicates that (1) there have been limited changes in non point
sources of nutrients and (2) that those changes have limited effect on 
water quality in Lavaca Bay to date. Although the study did not identify
strong responses to changes in nutrient loading, this does provide a baseline
assessment for future water quality management activities in the watershed.

>>>>

This study used an information theoretic approach to evaluate 
if nutrient loads and/or freshwater inflows provided evidence of effects on water quality concentrations in Lavaca Bay. Model probabilities were calculated and compared using the AIC~c~ scores between each group of temporal, flow, and flow+load models [@burnhamAICModelSelection2011]. Improvements in model probabilities provide evidence that the terms explain additional variation in the response variable. If model probabilities were tied, there wasn't evidence the more complicated model explains additional variation in water quality.

@@

Using evaluation criteria from @moriasiHydrologicWaterQuality2015, predictive 
performance of nutrient loading GAMs ranged from "satisfactory" to
"very good" based on median NSE, r^2^, and PBIAS metrics calculated using 5-fold 
cross validation. Median NSE values were 0.34 (Lavaca River) and 
0.48 (Navidad River) for NO~3~ loads and 0.81 (Lavaca River) and 0.91 (Navidad 
River) for TP loads. Median r^2^ values were 0.70
(Lavaca River) and 0.87 (Navidad River) for NO~3~ loads and 0.93 (Lavaca River) 
and 0.99 (Navidad River) fopr TP loads. Median PBIAS values were 2.00
(Lavaca River) and 10.90 (Navidad River) for NO~3~ loads and -7.20 
(Lavaca River) and -3.30 (Navidad River) for TP loads. Density plots of
metrics show similar distribution of values between sites for the 
same parameter, with the exception r^2^ values for NO~3~ loads where 
USGS-08164000 showed a much larger variance in values compared to USGS-08164525
(Figure \ref{fig:fig2}). TP GAMS had higher average NSE and r^2^ values and 
lower variance in metric values compared to NO~3~.

@@

Predicted annual NO~3~ and TP loads show considerable variation, generally following
patterns in discharge (Figure \ref{fig:fig3}, Figure \ref{fig:fig4}). 
Flow-normalized TP loads at both sites and flow-normalized NO~3~ loads at 
USGS-08164000 indicate watershed based loads did not change much over time
when accounting for variation driven by streamflow (Figure \ref{fig:fig3}). 
Flow-normalized loads at USGS-08164000 show small variation over time with 
some decreases in NO~3~ loads since 2013.

Aggregated across both sites, the mean 
annual NO~3~ load 2005 through 2020 was `r mean_no3` kg (`r mean_no3_low` kg - 
`r mean_no3_upper` kg, 90% CI). Annual NO~3~ loads ranged from `r min_no3` kg 
in `r min_no3_year` to `r max_no3` kg in `r max_no3_year`. Total annual TP loads 
ranged from `r min_tp` kg in `r min_tp_year` to 
`r max_tp` kg  in `r max_tp_year`. Mean annual TP loading from 2005 through 2020
was `r mean_tp` kg (`r mean_tp_low` kg - `r mean_tp_upper` kg, 90% CI).
On average, USGS-08164525 accounted for `r no3_mean_prop` of NO~3~ loads and 
`r tp_mean_prop` of TP loads from 2005 through 2020. However, during periods of extreme 
drought the Lavaca River (USGS-08164000) became the primary source of 
nutrient loading in the watershed with the Navidad River only accounting for 
`r min_no3_prop` and `r min_tp_prop` of NO~3~ and TP loads in 2011 (Figure 
\ref{fig:fig4}).

@@

Converted to average annual yield, the estimates of annual TP loads at
Lavaca River (USGS-08164000) are within the ranges in previous published studies
(Table \ref{tab:tpcomp}; [@dunnTrendsNutrientInflows1996; @rebich_sources_2011; @omaniEstimationSedimentNutrient2014; @wise_spatially_2019]). It isn't obvious
why TP estimates in @dunnTrendsNutrientInflows1996 were notably lower. Given that none
of the studies identify substantially sized trends in TP, it is is possible that 
the period used in @dunnTrendsNutrientInflows1996 was drier on average than the
other studies. The SPARROW models used in @rebich_sources_2011 and 
@wise_spatially_2019 utilize a version of LOADEST in the underlying load
estimation procedure, so a difference due to methodology alone is unlikely.

@@

Cross-validation of the GAM loading models indicated that GAMs performed well on
average at predicting daily nutrient loading values. The variance in scores was
very high indicating subsets of values were problematic at characterizing functional
relationships between nutrients and predictors. Because all of the water quality
data for these two locations in the TCEQ databases were ambient water quality
data, collected to be representative of typical flow conditions, there were few
data at the highest portions of the flow-duration curve. It was beyond the scope
of the current study to evaluate the subsets of cross-validation data and scores.
However, the cross-validation procedure is indicative that more robust sampling
would be beneficial for reducing prediction variance. Supplementary flow-biased 
monitoring targeting storm- or high-flow conditions is recommended here to
improve the precision of GAM predictions [@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017].

The non-linear temporal water quality trends identified using GAMs differed
slightly from the trends identified by @bugica_water_2020. This is
not unexpected due to the different time periods, different methodology, and
generally small slopes identified for most of the significant water quality
parameters in prior work. The trend in DO and 
cholorophyll-*a* concentrations are stable in comparison to other Texas
estuaries that are facing larger demands for freshwater diversions, higher
population growth, and more intense agricultural production 
[@wetzWaterQualityDynamics2016; @bugica_water_2020]. The trend of 
increasing NO~*x*~ concentration at the upper-bay TCEQ-13563 site and recent 
increases in TKN concentration at the mid-bay TCEQ-13383
site are concerning due to the nitrogen limitation identified in many Texas 
estuaries [@gardnerNitrogenFixationDissimilatory2006; 
@houTransformationFateNitrate2012; @doradoUnderstandingInteractionsFreshwater2015; 
@paudelRelationshipSuspendedSolids2019; @wetz_exceptionally_2017] 
and the relatively low ambient concentrations observed in Lavaca Bay.

The strong positive effect of freshwater inflow on NO~*x*~, TKN, and TP are
suggestive of nonpoint watershed sources, consistent with watershed uses and
with other studies relating freshwater inflow with nutrient 
concentrations in Lavaca Bay and other estuaries [@russell_effect_2006; 
@caffreyHighNutrientPulses2007; @peierlsNonmonotonicResponsesPhytoplankton2012; 
@palmerImpactsDroughtsLow2015; @ciraPhytoplanktonDynamicsLowInflow2021]. Inflow 
had a non-linear relationship with TKN at the two upstream sites, with TKN 
increasing as freshwater inflow transitioned from low to moderate levels. 
At higher freshwater inflows, the effect was attenuated, possibly indicating a 
flushing effect at higher freshwater inflow. No relationship between TKN and 
freshwater inflow were observed at TCEQ-13384 located in the lower reach of 
Lavaca Bay. Tidal flushing from Matagorda Bay could be responsible for 
diluting TKN and acting as a control on the effects of freshwater inflow in 
lower reaches of Lavaca Bay. @russell_effect_2006 also suggested the processing of 
organic loads in the upper portions of Lavaca Bay reduces the transport of 
nutrients into the lower reaches of the Bay.

Freshwater inflow had a strong positive effect on chlorophyll-*a* at 
the upper- and mid-bay sites. The upper-bay site, TCEQ-13563, showed decreases
in chlorophyll-*a* at the highest freshwater inflow volumes. Freshwater flushing
or increases in turbidity are associated with decreases in chlorophyll-*a* in
other estuaries [@peierlsNonmonotonicResponsesPhytoplankton2012; 
@cloernPhytoplanktonPrimaryProduction2014]. No relationships between inorganic
nitrogen or TP loadings with chlorophyll-*a* were observed. Due to the lack
of TKN loading information, no assessment between organic nitrogen loads and
chlorophyll-*a* were possible.

Although other studies have identified complex relationships between estuary 
nutrient concentrations, nutrient loading and chlorophyll-*a* concentrations in 
Texas estuaries [@ornolfsdottirNutrientPulsingRegulator2004; 
@doradoUnderstandingInteractionsFreshwater2015; @ciraPhytoplanktonDynamicsLowInflow2021; @tominackVariabilityPhytoplanktonBiomass2022], 
this study specifically used flow-adjusted freshwater derived nutrient loads to 
parse out contributions from changes in nutrient loadings while accounting for 
variations in load due to flow. Loading GAMs indicated no evidence of changes in 
flow-normalized TP loads in either river, and no changes in flow-normalized NO~3~
loads in the Navidad River. The small changes in flow-normalized NO~3~ loads in
the Lavaca River are probably masked under most conditions by discharge from the
Navidad River. Given the relatively small variation in flow-normalized loads,
it can be expected that they would contribute little to the variance in downstream
water quality.

GAMs did not identify responses in DO concentration to inflows or nutrient loads.
The seasonality term in the temporal GAM models explained a substantial amount of DO
variation at all of the sites. Responses of estuary metabolic processes and 
resulting DO concentrations can be quite complicated and often locally specific [@caffreyFactorsControllingNet2004]. While the lack of total nitrogen or TKN 
loading data hinders interpretation, the large seasonal effect on DO suggests 
physical factors play an important role and should be included in future models.
Prior work suggests that Lavaca Bay may not be limited by nutrients alone, 
with high turbidity or nutrient processing in upper portions of the Bay or 
intertidal river limiting production [@russell_effect_2006]. Finally, it
is reasonable to assume that fluctuations in DO may not occur immediately in 
response to nutrient pulses or freshwater inflow. Work has has shown that various
water quality parameters may have lagged effects lasting days or even months following
storms and large discharge events [@mooney_watershed_2012; @wetzExtremeFutureEstuaries2013; @bukaveckas_influence_2020; @walkerTimescalesMagnitudeWater2021]. 
However, our work only evaluates responses to
loading and inflows occuring the day of water quality observations.

@@

GAM models appear to provide reliable estimates of nutrient loads in the Lavaca
Bay watershed. However, additional flow-biased data collection efforts would
decrease the prediction variance and improve accuracy at critical 
high flow events. Ongoing projects will fill data gaps for total nitrogen and 
TKN loading. This study, consistent with others along the Texas coast, found
strong effects of freshwater flow on nutrient and chlorophyll-*a* 
concentrations. DO concentrations, dominated by seasonal effects, did not show
strong direct responses to freshwater flow. Small variance in flow-adjusted
nutrient loads indicates that (1) there have been limited changes in non-point
sources of nutrients and (2) that there isn't strong evidence that those small 
changes have had effects on chlorophyll-*a* or dissolved oxygen in Lavaca Bay. 
Although the study did not identify strong responses to changes in nutrient 
loading, this does provide a baseline assessment for future water quality 
management activities in the watershed.
